---
title: "Project2"
author: "Nicholas Kunze"
date: "2024-03-24"
output:
  pdf_document: default
  html_document:
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
params:
  dbuser: 
    label: "Username"
    value: "nicholas.kunze77"
    input: text
  dbpass: 
    label: "Password"
    value: "nicholas.kunze77"
    input: password
  dbname: 
    label: "Database"
    value: "nicholas.kunze77"
    input: password
  dbhost: 
    label: "Host"
    value: "cunydata607sql.mysql.database.azure.com"
    input: password
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if("tidyverse" %in% rownames(installed.packages()) == FALSE) {install.packages("tidyverse")}
if("RMySQL" %in% rownames(installed.packages()) == FALSE) {install.packages("RMySQL")}
library(RMySQL)
library(tidyverse)
library(dplyr)
library(stringr)
```

## Summary

In this project, I showcase preparing datasets for downstream analysis work; this specifically deals with wide datasets and getting them ready by making them follow the rules of a tidy dataset. For this project, I chose my own discussion item, world population data and projects. 

## World Population History and Projections

Get table from MySQL server and get head of new DataFrame.

```{r getdb-pop}
azuredb = dbConnect(MySQL(), user=params$dbuser, password=params$dbpass, dbname=params$dbname, host=params$dbhost)

pop <- dbGetQuery(azuredb, 
  "SELECT 
      *
   FROM
      world_population_data;")
head(pop)
```

As we can see, this is a wide df showing information for a single country over multiple years with a few unneeded columns. We'd like each row to be a single observation of population for a country. We should gather up these years that are columns, as they are values for 'year', not variable names. Then we can just select the variables we need for analysis.

```{r pop-pivot-years}
pop <- pop %>% 
  pivot_longer(cols = `1980`:`2050`, names_to = "year", values_to = "population")
pop <- pop[,c("country","year","population","landAreaKm","growthRate")]
head(pop)
```

```{r pop-analyze-top-5}
pop_top <- head(pop,30)
ggplot(data = pop_top, mapping = aes(x = landAreaKm, y = population, color = country)) +
  geom_point()
```
It looks like our big 5 nations are actually slowing down over time regarding their pop growth.

Now, I'm curious if density has any correlation with growth...

```{r pop-analyze-growth-density}
pop_density <- mutate(pop, density = population/landAreaKm)
cor(pop_density$density, pop_density$growthRate)
ggplot(data = pop_density, mapping = aes(x = density, y = growthRate, color = country)) +
  geom_point(show.legend = FALSE)
```
It does appear that there is a decent enough negative correlation between population density and growth rate.
